---
title: "Prestige and Placement in North American Philosophy"
author: "Jonathan Weisberg"
output:
  md_document:
    variant: markdown
css: custom.css
---

How does prestige correlate with placement in academic philosophy? There's good stuff on this already, like [this post](http://placementdata.com:8182/the-philosophical-gourmet-report-and-placement/) by Carolyn Dicey Jennings, Pablo Contreras Kallens, and Justin Vlasits.[^0] This post uses the same data sources, but emphasizes different things (visualization, North American PhDs, and primarily tenure-track jobs).

[^0]: Also check out Figure 1 in [this paper](https://quod.lib.umich.edu/e/ergo/12405314.0005.010?view=text;rgn=main) by Helen De Cruz.

```{r setup, include=FALSE, warning=FALSE}
library(tidyverse)
library(ggrepel)
library(hrbrthemes)

my_theme <- list(
  labs(caption ="Data sources: www.placementdata.com and www.philosophicalgourmet.com\nAuthor: Jonathan Weisberg (@jweisber)"),
  theme_ipsum() 
)

# load data ----
df_pgr <- read_csv("data/pgr.csv")
positions <- read_csv("data/Position_view.csv")
universities <- read_csv("data/University_view.csv")
graduations <- read_csv("data/Graduation_view.csv") %>% select(-id)
placements <- read_csv("data/Placement_view.csv", 
                       col_types = list(university_id = col_character())) %>% select(-id)

graduations <- graduations %>%
  rename(program_graduated_id = university, year_graduated = year)

# lump HPS programs
graduations$program_graduated_id[graduations$program_graduated_id == 500] <- 499
graduations$program_graduated_id[graduations$program_graduated_id == 557] <- 12
graduations$program_graduated_id[graduations$program_graduated_id == 556] <- 375
graduations$program_graduated_id[graduations$program_graduated_id == 559] <- 19
graduations$program_graduated_id[graduations$program_graduated_id == 5413] <- 132
placements$university_id[placements$university_id == 500] <- 499
placements$university_id[placements$university_id == 557] <- 12
placements$university_id[placements$university_id == 556] <- 375
placements$university_id[placements$university_id == 559] <- 19
placements$university_id[placements$university_id == 5413] <- 132

placements <- placements %>%
  rename(program_placed = university, year_placed = year)

placements$university_id[placements$university_id == "NULL"] <- NA
placements$university_id <- as.integer(placements$university_id)

placements$year_placed[placements$year_placed == 19182] <- 1982
placements$year_placed[placements$year_placed == 2104] <- 2014

universities$name <- gsub("\\*", "", universities$name)
universities$name[universities$name == "King´s College London"] <- "King's College London"
universities$name[universities$name == "Texas A & M University-College Station"] <- "Texas A&M University"
universities$name[universities$name == "University of Southampton"] <- "University of Southampton"
universities$name[universities$name == "Unviersity of Birmingham"] <- "University of Birmingham"
universities$name[universities$name == "University of Canterbury Christchurch"] <- "University of Canterbury"
universities$name[universities$name == "Queen´s University"] <- "Queen's University"

df_apda <- graduations %>%
  left_join(placements, by = "person") %>%
  left_join(universities, by =  c("program_graduated_id" = "id")) %>%
  rename("program_graduated" = "name") %>%
  left_join(positions, by = c("position" = "id"))

df_apda$title[is.na(df_apda$title)] <- "Unknown"

df_pgr_nam_2006 <- df_pgr %>% filter(year == 2006, locale %in% c("Canada", "US"))

pgr_ids <- universities %>% filter(name %in% df_pgr_nam_2006$program) %>% .$id
phd_ids <- graduations$program_graduated_id %>% unique()

df <- df_apda %>% 
  filter(year_graduated %in% seq(2012, 2014)) %>%
  inner_join(df_pgr_nam_2006, by = c("program_graduated" = "program"))
```


# TT Placement in North America

Let's start with a simple question of broad interest. In North America, how well does the [PGR](https://www.philosophicalgourmet.com/) rating of one's PhD-granting program predict one's chances of landing a tenure-track (TT) job?

Consider all the people who got a PhD from a North American philosophy program in the years 2012--14.[^1] Focus for now on those from PhD programs ranked by the 2006 edition of the PGR.[^2] Now group them according to those PGR ratings, rounded to the nearest 0.5.

[^1]: Why these years? Because that's where the data is best. The APDA has focused its collection efforts so far on graduates from the years 2012--16, so that's where the data is the most plentiful. But the data for 2015 and 2016 graduates probably aren't "ripe" enough yet for our purposes; many graduates who will ultimately find TT jobs are probably still in postdocs and other temporary gigs. Thanks to Brian Weatherson for [pushing me to take this into account](https://twitter.com/bweatherson/status/1134570972315508736).

    Of course, the 2012--2014 data aren't fully ripe either. But [previous noodling](https://jonathanweisberg.org/post/prestige-and-time-to-tt/) suggests they're probably pretty close.

[^2]: Why the 2006 edition? Partly for continuity with [the APDA's own analysis](http://placementdata.com:8182/the-philosophical-gourmet-report-and-placement/). But also because students often use PGR rankings to choose PhD programs, and the rankings available to them typically predate the year of their PhD by 6 or 7 years.

This gives us 7 groups of PhDs (rankings range from 2.0 to 5.0). According to [the APDA's data](http://placementdata.com), the portion from each group who ended up in TT jobs are as follows:

```{r echo=FALSE, dpi=300}
df1 <- df %>%
  group_by(person) %>%
  mutate(postdoc = any(title == "Fellowship/Postdoc"),
         perm = any(title %in% c("Tenure-Track", "Lecturer (Permanent)", "Instructor (Permanent)", 
                                 "Adjunct (Permanent)", "Other (Permanent)")),
         tt = any(title == "Tenure-Track"),
         phd = any(title == "Tenure-Track" & university_id %in% phd_ids),
         pgr = any(title == "Tenure-Track" & university_id %in% pgr_ids)) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(mean_rounded = round(mean / .5) *.5) %>%
  group_by(mean_rounded) %>%
  summarize(n = n(), 
            postdoc_rate = sum(postdoc) / n,
            perm_rate = sum(perm) / n,
            tt_rate = sum(tt) / n,
            phd_rate = sum(phd) / n,
            pgr_rate = sum(pgr) / n) %>%
  gather(key = "rate_type", value = "rate", postdoc_rate, perm_rate, tt_rate, phd_rate, pgr_rate)
  
ggplot(df1 %>% filter(rate_type == "tt_rate"), aes(mean_rounded, rate)) +
  geom_line() +
  lims(x = c(2,5), y = c(0,1)) +
  labs(x = "PGR Rating (rounded to nearest 0.5)",
       y = "TT Placement Rate",
       title = "Prestige & TT Placement in North American Philosophy",
       subtitle = "2012-14 Philosophy PhDs, grouped by the 2006 PGR Rating of their PhD-granting program") +
  my_theme
```

There's clearly a positive connection; almost perfectly linear in fact. And the gist---*very crudely speaking*---is that a high prestige PhD about doubles your chances of landing a TT job over a low-prestige PhD: from ~30% to ~60%.

Note that the data are sparse at the extremes though. Consider this raw look, where each point is a PhD graduate.

```{r echo=FALSE, dpi=300}
df2 <- df %>%
  group_by(person) %>%
  mutate(tt = ifelse(any(title == "Tenure-Track"), "TT Job", "No TT Job")) %>%
  slice(1)

ggplot(df2, aes(mean, tt)) +
  geom_jitter(aes(colour = tt), height = 0.1, width = 0.1, alpha = 0.3) +
  scale_colour_brewer(palette = "Set1") +
  labs(x = "PGR Rating",
       y = "",
       title = "Prestige & TT Placement in North American Philosophy",
       subtitle = "TT Placements by PGR Rating of PhD-granting Program (2006 PGR, 2012-14 PhDs)") +
  my_theme +
  theme(legend.position = "none")
```

A "violin plot" shows the same thing but easier to read: the thickness of the violins indicates the density of points at each *x*-position.

```{r echo=FALSE, dpi=300}
ggplot(df2, aes(tt, mean)) +
  geom_violin(aes(fill = tt)) +
  scale_fill_brewer(palette = "Set1") +
  coord_flip() +
  labs(x = "",
       y = "PGR Rating",
       title = "Prestige & TT Placement in North American Philosophy",
       subtitle = "TT Placements by PGR Rating of PhD-granting Program (2006 PGR, 2012-14 PhDs)") +
  my_theme +
  theme(legend.position = "none")
```

With so few points at the ends, we shouldn't read too much into the exact placement rates there.


# Other Placement Types

What about other kinds of jobs? Let's consider five categories, defined as follows.

- Postdoc: "Fellowship/Postdoc" in the APDA database.
- Permanent: any of the following in the APDA database.
  - "Tenure-Track"
  - "Lecturer (Permanent)"
  - "Instructor (Permanent)"
  - "Adjunct (Permanent)"
  - "Other (Permanent)"
- Tenure-Track: "Tenure-Track" in the APDA database.
- PhD Program: Tenure-Track at a PhD-granting program.
- PGR Ranked: Tenure-Track at a 2006 PGR-ranked program.

These are hardly perfect definitions, but they're manageable with this data while still being pretty informative.

```{r echo=FALSE, dpi=300}
ggplot(df1, aes(mean_rounded, rate)) +
  geom_line(aes(colour = rate_type)) +
  scale_colour_brewer(palette = "Set1", name = "Type",
                      breaks = c("postdoc_rate", "perm_rate", "tt_rate", "phd_rate", "pgr_rate"),
                      labels = c("Postdoc", "Permanent", "Tenure-Track", "PhD Program", "PGR Ranked")) +
  lims(x = c(2,5), y = c(0,1)) +
  labs(x = "PGR Rating (rounded to nearest 0.5)",
       y = "Placement Rate",
       title = "Prestige & Placement in North American Philosophy",
       subtitle = "2012-14 Philosophy PhDs, grouped by the 2006 PGR Rating of their PhD-granting program") +
  my_theme
```

Note that a graduate can appear in multiple categories (Tenure-Track is a subset of Permanent, after all).


# Unranked Programs

What about PhD programs not ranked in the 2006 PGR?[^3] The numbers may be iffier here. Some programs have only one graduate listed for example, a graduate who got a TT job. But there are only a few such programs, and more than 600 graduates otherwise. So the numbers may still be good approximations.

[^3]: Thanks to [Amanda at the Philosophers' Cocoon](https://philosopherscocoon.typepad.com/blog/2019/05/phd-program-prestige-and-tt-placement.html?cid=6a014e89cbe0fd970d0240a48b2f6b200d#comment-6a014e89cbe0fd970d0240a48b2f6b200d) for prompting me to look at this.

```{r echo=FALSE}
df3 <- df_apda %>% 
  filter(year_graduated %in% seq(2012, 2014),
         country %in% c("U.S.", "Canada"),
         !(program_graduated %in% df_pgr_nam_2006$program),) %>%
  group_by(person) %>%
  mutate(postdoc = any(title == "Fellowship/Postdoc"),
         perm = any(title %in% c("Tenure-Track", "Lecturer (Permanent)", "Instructor (Permanent)", 
                                 "Adjunct (Permanent)", "Other (Permanent)")),
         tt = any(title == "Tenure-Track"),
         phd = any(title == "Tenure-Track" & university_id %in% phd_ids),
         pgr = any(title == "Tenure-Track" & university_id %in% pgr_ids)) %>%
  slice(1) %>%
  ungroup()

df4 <- df3 %>%
  summarize(postdoc_rate = round(sum(postdoc) / n(), 2),
            perm_rate = round(sum(perm) / n(), 2),
            tt_rate = round(sum(tt) / n(), 2),
            phd_rate = round(sum(phd) / n(), 2),
            pgr_rate = round(sum(pgr) / n(), 2))

knitr::kable(df4, col.names = c("Postdoc", "Permanent", "TT", "PhD", "PGR"), format = "pandoc")
```

If you're curious which programs stand out among the unranked, here are the top 10 by TT placement (excluding those with 5 or fewer graduates).

```{r echo=FALSE}
df5 <- df3 %>%
  group_by(program_graduated) %>%
  summarize(n = n(), 
            postdoc_rate = round(sum(postdoc) / n(), 2),
            perm_rate = round(sum(perm) / n(), 2),
            tt_rate = round(sum(tt) / n(), 2),
            phd_rate = round(sum(phd) / n(), 2),
            pgr_rate = round(sum(pgr) / n(), 2))

knitr::kable(
  df5 %>% 
    filter(n > 5) %>% 
    top_n(10, tt_rate) %>%
    arrange(desc(tt_rate)),
  col.names = c("Program", "N", "Postdoc", "Permanent", "TT", "PhD", "PGR"),
  format = "pandoc"
)
```

Note that the top 3 are at Christian universities, and as you might expect, a lot of their placement is driven by hires at Christian schools.

Here are the 10 "largest" programs, i.e. those with the most graduates listed in the APDA database.

```{r echo=FALSE}
knitr::kable(
  df5 %>% 
    top_n(10, n) %>%
    arrange(desc(n)),
  col.names = c("Program", "N", "Postdoc", "Permanent", "TT", "PhD", "PGR"),
  format = "pandoc"
)
```


# Departmental TT Placement

Looking at placement rates by department raises the question: how well does a department's PGR rating predict its TT placement rate?

```{r echo=FALSE, dpi=300}
df6 <- df %>%
  group_by(program_graduated, person) %>%
  mutate(tt = any(title == "Tenure-Track")) %>%
  ungroup() %>%
  group_by(program_graduated) %>%
  mutate(n = n(), rate = sum(tt == TRUE) / n) %>%
  slice(1)

m <- lm(rate ~ mean, df6, weights = n)
df6$predicted_rate <- predict(m, df6)

shorten_names <- function(df, long, short) {
  for(i in 1:length(long)) {
    long_name <- long[i]
    short_name <- short[i]
    df$program_graduated[df$program_graduated == long_name] <- short_name
  }
  return(df)
}

df6 <- shorten_names(df6,
        c("University of Connecticut", "University of California, Riverside", 
          "University of Maryland, College Park", "McGill University", "Western University",
          "University of Massachusetts Amherst", "Ohio State University", "University of California, Davis",
          "University of Alberta", "University of British Columbia", "Syracuse University",
          "Massachusetts Institute of Technology", "Johns Hopkins University", "University of Pittsburgh"),
        c("Connecticut", "UC Riverside", "Maryland", "McGill", "Western", "Amherst", "Ohio",
          "UC Davis", "Alberta", "UBC", "Syracuse", "MIT", "Johns Hopkins", "Pittsburgh"))

ggplot(df6, aes(mean, rate)) +
  geom_point(aes(size = n)) +
  geom_abline(intercept = m$coefficients[1], slope = m$coefficients[2], colour = "blue") +
  labs(x = "PGR Rating",
       y = "TT Placement Rate",
       size = "Graduates",
       title = "Prestige & TT Placement in North American Philosophy",
       subtitle = "Departmental TT placement rates for 2012-14 graduates vs. 2006 PGR Rating",
       caption = "Data sources: www.placementdata.com and www.philosophicalgourmet.com\nAuthor: Jonathan Weisberg (@jweisber)") +
  my_theme
```

There's a clear connection, but also a lot of variation. Which are the programs that especially stand out from the trend? Suppressing the sizing for visibility, we can label those programs above/below the trendline by at least 0.2.

```{r echo=FALSE, dpi=300}
ggplot(df6, aes(mean, rate)) +
  geom_point() +
  geom_text_repel(data = df6 %>% filter(rate <= predicted_rate - .2 | rate >= predicted_rate + .2), 
                  aes(label = program_graduated)) +
  geom_abline(intercept = m$coefficients[1], slope = m$coefficients[2], colour = "blue") +
  labs(x = "PGR Rating",
       y = "TT Placement Rate",
       size = "Graduates",
       title = "Prestige & TT Placement in North American Philosophy",
       subtitle = "Departmental TT placement rates for 2012-14 graduates vs. 2006 PGR Rating",
       caption = "Data sources: www.placementdata.com and www.philosophicalgourmet.com\nAuthor: Jonathan Weisberg (@jweisber)") +
  my_theme
```

For complete listings of departmental placement rates, check out the APDA's infograms [here](http://www.placementdata.com/data/).

# Notes
